---
title: "Case study miRetrieve"
output: html_notebook
editor_options: 
  chunk_output_type: console
---
# Install packages if necessary
```{r}
install.packages(c("miRetrieve",
                   "magrittr",
                   "ggplot2",
                   "easyPubMed",
                   "dplyr",
                   "tidyr",
                   "patchwork"))
```

# Load libraries

```{r}
library(miRetrieve)
library(magrittr) # Load magrittr for %>%
library(ggplot2) # Load ggplot2 for plotting
library(easyPubMed) # Load easyPubMed, access to PubMed
library(dplyr) # Data wrangling
library(tidyr) # Data wrangling
library(patchwork) # Group graphs

# Set working directory with `setwd()` to folder containing 
# PubMed abstracts ("miRNA_atherosclerosis.txt", "miRNA_lungcancer.txt") and
# 200 randomly chosen PubMed IDs for miRNA extraction accuracy 
# ("PMIDs_miRNA_extraction.R")
# All data can be downloaded from
# https://github.com/JulFriedrich/miRetrieve/releases/tag/v1.3.2
```


# Load abstracts atherosclerosis

```{r}
# PubMed file of abstracts matching "atherosclerosis miRNA"
file <- "atherosclerosis_2021.txt"

# Load abstracts
df_atherosclerosis <- read_pubmed(file, topic = "Atherosclerosis") %>% 
    # Subset for original articles
    subset_research() %>%
    # Extract miRNA names mentioned at least twice/abstract
    extract_mir_df(threshold = 2)
    
# Number of abstracts
length(unique(df_atherosclerosis[["PMID"]]))

# Number of rows
nrow(df_atherosclerosis)
```

# Compare extraction count to PubMed

```{r}
# Extract all miRNAs
df_athero_comparison <- read_pubmed(file,
                        topic = "Atherosclerosis") %>% 
    extract_mir_df(threshold = 1,
                   extract_letters = TRUE)

# Count miRNAs
df_count_letter <- df_athero_comparison %>% 
    count_mir() %>% 
  # Filter for miRNAs with lettered suffixes
    dplyr::filter(stringr::str_detect(miRNA, "\\d[a-z]")) %>% 
    dplyr::slice(1:30)

# Filter for miRNAs with lettered suffixes
# Extract miRNA numbers of miRNAs with lettered suffixes
df_mir_letter <- df_athero_comparison %>% 
    count_mir() %>% 
    dplyr::filter(stringr::str_detect(miRNA, "\\d[a-z]")) %>% 
    filter(Mentioned_n >= 10) %>% 
    pull(miRNA) %>% 
    stringr::str_extract_all("miR-\\d+|let-7") %>% 
    unlist()

# Filter for miRNAs without lettered suffixes
df_count_no_letter <- df_athero_comparison %>% 
    count_mir() %>% 
    dplyr::filter(stringr::str_detect(miRNA, "\\d$")) %>% 
    filter(!miRNA %in% df_mir_letter) %>% 
    dplyr::slice(1:30)
```

## Obtain PubMed IDs

```{r}
# Obtain PubMed IDs from query "{miRNA} + atherosclerosis" 
# as of February 11th, 2021
compare_count <- function(mir) {
    query <- paste('(',mir, 'atherosclerosis) AND (("1990/01/01"[Date - Entry] : "2021/02/11"[Date - Entry]))')
    pubmed_ids <- get_pubmed_ids(query) %>% 
        fetch_all_pubmed_ids()
    return(pubmed_ids)
}

count_pubmed_vec <- purrr::map(df_count_no_letter$miRNA, compare_count) %>% 
    purrr::set_names(df_count_no_letter$miRNA)

count_pubmed_vec_letter <- purrr::map(df_count_letter$miRNA, compare_count) %>% 
    purrr::set_names(df_count_letter$miRNA)
```

## Compare miRetrieve with PubMed's miRNA count

```{r}
# Obtain number of PubMed results
length_pmid <- purrr::map(count_pubmed_vec, length) %>% 
    unlist()

# Add number of PubMed results to dataframe "df_count_no_letter"
df_count_no_letter[3] <- length_pmid
names(df_count_no_letter)[3] <- "PubMed"
names(df_count_no_letter)[2] <- "miRetrieve"

# Plot # of miRNAs with miRetrieve vs. PubMed (no letters)
count_plot <- df_count_no_letter %>%
    mutate(miRNA = forcats::fct_reorder(miRNA, miRetrieve)) %>% 
    pivot_longer(cols = c(miRetrieve, PubMed)) %>%
    mutate(name = forcats::fct_rev(name)) %>% 
    ggplot(aes(x = miRNA, y = value, fill = name)) +
    geom_col(position = "dodge") +
    coord_flip() +
    guides(fill = FALSE) +
    scale_fill_brewer(palette = "Dark2") +
    theme_classic() +
    ylab("# of articles") + 
    scale_y_continuous(expand = c(0,0))

# Obtain number of PubMed results
length_pmid_letter <- purrr::map(count_pubmed_vec_letter, length) %>% 
    unlist()

# Add number of PubMed results to dataframe "df_count_letter"
df_count_letter[3] <- length_pmid_letter
names(df_count_letter)[3] <- "PubMed"
names(df_count_letter)[2] <- "miRetrieve" 

# Plot # of miRNAs with miRetrieve vs. PubMed (with letters)
count_letter_plot <- df_count_letter %>%
    mutate(miRNA = forcats::fct_reorder(miRNA, miRetrieve)) %>% 
    pivot_longer(cols = c(miRetrieve, PubMed)) %>%
    mutate(name = forcats::fct_rev(name)) %>% 
    ggplot(aes(x = miRNA, y = value, fill = name)) +
    geom_col(position = "dodge") +
    coord_flip() +
    guides(fill = guide_legend(reverse = TRUE, title = "Method")) +
    scale_fill_brewer(palette = "Dark2") +
    theme_classic()+
    ylab("# of articles")+ 
    scale_y_continuous(expand = c(0,0))

# Combine plots (Figure 1)
combined <- count_plot + count_letter_plot & theme(legend.position = "bottom")

combined + 
    plot_layout(guides = "collect") + 
    plot_annotation(tag_levels = "A")

# Calculate score [miRetrieve count] / [PubMed count] (no letters)
df_count_no_letter %>% 
    mutate(Score = miRetrieve / PubMed) %>%
    mutate(Score = ifelse(is.infinite(Score), 0.01, Score)) %>%
    summarise(mean(Score), sd(Score), sum(miRetrieve), sum(PubMed))

# Calculate score [miRetrieve count] / [PubMed count] (with letters)
df_count_letter %>% 
    mutate(Score = miRetrieve / PubMed) %>%
    mutate(Score = ifelse(is.infinite(Score), 0.01, Score)) %>%
    summarise(mean(Score), sd(Score), sum(miRetrieve), sum(PubMed))
```

# Validate miRNA extraction accuracy

```{r}
load("PMIDs_miRNA_extraction.R")

# Extract miRNAs mentioned at least once per abstract (no letters)
df_atherosclerosis_test_extraction <- read_pubmed(file,
                                                  topic = "Atherosclerosis") %>%
    subset_research() %>% 
    extract_mir_df(threshold = 1) %>%
    dplyr::group_by(PMID, Year, Title, Abstract) %>%
    dplyr::summarise(miRNA_coll = paste(miRNA, collapse = ", "), .groups = "rowwise" ) %>%
    dplyr::ungroup()

pmids_test %>% 
  left_join(df_atherosclerosis_test_extraction,
            by = "PMID") %>% 
  save_excel(excel_file = "miRetrieve_extraction.xlsx")

# Extract miRNAs mentioned at least once per abstract (with letters)
df_atherosclerosis_test_extraction_letter <- read_pubmed(file,
                                                     topic = "Atherosclerosis") %>%
    subset_research() %>% 
    extract_mir_df(threshold = 1, extract_letters = TRUE) %>%
    dplyr::group_by(PMID, Year, Title, Abstract) %>%

    dplyr::summarise(miRNA_coll = paste(miRNA, collapse = ", "), .groups = "rowwise") %>% 
    dplyr::ungroup()

pmids_test %>% 
  left_join(df_atherosclerosis_test_extraction_letter,
            by = "PMID") %>% 
  save_excel(excel_file = "miRetrieve_extraction_letter.xlsx")
```

# Top miRNAs in atherosclerosis

```{r}
# Plot top miRNAs (Figure 2)
plot_mir_count(df_atherosclerosis,
               title = "Most frequently mentioned miRNAs in atherosclerosis")
```

# Stopwords atherosclerosis

```{r}
# Generate new stop word list specifically for atherosclerosis to 
# improve overview
atherosclerosis_stop <- generate_stopwords(c("atherosclerosis",
                                           "atherosclerotic",
                                           "patients",
                                           "pathogenesis",
                                           "umbilical",
                                           "huvecs",
                                           "umbilical vein",
                                           "human umbilical",
                                           "cardiovascular disease",
                                           "atherosclerosis however",
                                           "atherosclerosis as",
                                           "development atherosclerosis",
                                           "pathogenesis atherosclerosis",
                                           "oxidized low",
                                           "vein endothelial",
                                           "cardiovascular diseases",
                                           "cells huvecs",
                                           "necrosis factor",
                                           "cells vsmcs",
                                           "cells ecs",
                                           "demonstrated mir",
                                           "low density",
                                           "atherosclerotic plaque",
                                           "artery disease",
                                           "muscle cells",
                                           "tumor necrosis",
                                           "coronary artery",
                                           "dependent manner",
                                           "vascular endothelial",
                                           "patients coronary",
                                           "serum levels",
                                           "vascular smooth",
                                           "α"),
                                           combine_with = stopwords_miretrieve)
```

# Top term atherosclerosis

```{r}
# Plot single word terms for miR-155
mir155_1 <- plot_mir_terms(df_atherosclerosis, "miR-155", 
                           stopwords = atherosclerosis_stop,
                           top = 15,
                           title = "Top single terms for miR-155 in atherosclerosis") +
  theme(axis.text.y = element_text(size = 12),
        plot.title = element_text(size = 10),
        axis.title = element_text(size = 10))

# Plot 2-grams for miR-155
mir155_2 <-  plot_mir_terms(df_atherosclerosis, "miR-155", token = "ngrams", n = 2,
               stopwords = atherosclerosis_stop,
               top = 14,
               title = "Top 2-grams for miR-155 in atherosclerosis") +
  theme(axis.text.y = element_text(size = 12),
        plot.title = element_text(size = 10),
        axis.title = element_text(size = 10))

# Plot single word terms for miR-21
mir21_1 <-  plot_mir_terms(df_atherosclerosis, "miR-21",
               stopwords = atherosclerosis_stop,
               top = 11,
               title = "Top single terms for miR-21 in atherosclerosis") +
  theme(axis.text.y = element_text(size = 12),
        plot.title = element_text(size = 10),
        axis.title = element_text(size = 10))

# Plot 2-grams for miR-21
mir21_2 <- plot_mir_terms(df_atherosclerosis, "miR-21", token = "ngrams", n = 2,
               stopwords = atherosclerosis_stop,
               top = 13,
               title = "Top 2-grams for miR-21 in atherosclerosis") +
  theme(axis.text.y = element_text(size = 12),
        plot.title = element_text(size = 10),
        axis.title = element_text(size = 10))

# Combine plots (Figure 3)
(mir155_1 | mir155_2) /
    (mir21_1 | mir21_2) + plot_annotation(tag_levels = 'A')
```

# Targets in atherosclerosis

```{r}
# Add miRTarBase targets to df
df_targets_ath_mir <- join_mirtarbase(df_atherosclerosis)

# Plot targets of miR-155 and miR-21 (Figure 4)
plot_target_mir_scatter(df_targets_ath_mir,
                        mir = c("miR-155", "miR-21"),
                        alpha = 1,
                        title = "Targets of miR-21 and miR-155 in atherosclerosis")
```

# Lung cancer abstracts loading

```{r}
# PubMed file of abstracts matching "lung cancer miRNA"
file_lc <- "lungcancer_2021.txt"

# Load abstracts
df_lc <- read_pubmed(file_lc, topic = "Lung cancer") %>% 
    # Subset for original articles
    subset_research() %>%
    # Extract miRNA names mentioned at least twice/abstract
    extract_mir_df(threshold = 2)

# Number of abstracts
length(unique(df_lc[["PMID"]]))

# Number of rows
nrow(df_lc)
```

# Top Terms Lung cancer

```{r}
# Plot top miRNAs in lung cancer (Figure 5)
plot_mir_count(df_lc,
               title = "Most frequently mentioned miRNAs in lung cancer")
```

# Compare miR-21-term associations in atherosclerosis and lung cancer 

```{r}
# Generate stop words for miR-21 comparison to improve overview
comparison_stop <- generate_stopwords(c("dependent manner",
                                        "vascular endothelial",
                                        "peripheral blood",
                                        "serum mir",
                                        "serum levels",
                                        "tensin homolog",
                                        "phosphatase tensin",
                                        "α"),
                                      combine_with = stopwords_miretrieve)
```

```{r}
# Combine `df_atherosclerosis` and `df_lc`
df_athlc <- combine_df(df_atherosclerosis, df_lc)

# Compare shared terms, single word
comp_mir21_1 <- compare_mir_terms(df_athlc, "miR-21", top = 7,
                                  title = "Comparison of miR-21 single term association",
                                  stopwords = comparison_stop) +
  theme(axis.text.y = element_text(size = 12),
        plot.title = element_text(size = 10),
        axis.title = element_text(size = 10)) +
  ggplot2::scale_fill_manual(values = c("#F5793A", "#85C0F9"), name = "Topic")

# Compare shared terms, 2-gram
comp_mir21_2 <- compare_mir_terms(df_athlc, "miR-21", token = "ngrams", n = 2, top = 9,
                                  title = "Comparison of miR-21 2-gram association",
                                  stopwords = comparison_stop) +
  theme(axis.text.y = element_text(size = 12),
        plot.title = element_text(size = 10),
        axis.title = element_text(size = 10)) +
  ggplot2::scale_fill_manual(values = c("#F5793A", "#85C0F9"), name = "Topic")

# Combine plot (Figure 6)
plot_combined <- (comp_mir21_1 | comp_mir21_2) & theme(legend.position = "bottom")

plot_combined + 
    plot_annotation(tag_levels = 'A')  + 
    plot_layout(guides = "collect")
```

# Potential miRNA biomarker in atherosclerosis and lung cancer

```{r}
# Potential miRNA biomarker in atherosclerosis
ath_biomarker <- calculate_score_biomarker(df_atherosclerosis,
                                           threshold = 5,
                                           discard = TRUE)


# Plot top 7 miRNA biomarker in atherosclerosis
bm_ath_plot <- plot_mir_count(ath_biomarker, 
                              top = 8,
                              title = "Potential biomarker miRNAs in atherosclerosis")

# Potential miRNA biomarker in lung cancer
lc_biomarker <- calculate_score_biomarker(df_lc,
                                           threshold = 5,
                                           discard = TRUE)

# Plot top 7 miRNA biomarker in lung cancer
bm_lc_plot <- plot_mir_count(lc_biomarker, 
                             top = 8,
                             title = "Potential biomarker miRNAs in lung cancer")

# Combine plot (Figure 7)
(bm_ath_plot) + (bm_lc_plot) + plot_annotation(tag_levels = 'A')
```
